A sample workflow

### 2025-04-22 analyses/african_mammals_workflow.qmd
start working on workflow with African mammals

# A sample workflow

@@

@@

@@

What is trend in terrestrial vertebrate species population numbers in South Africa, both overall and how does it vary acrosss species and populations?

Now we start to simulate! 
*Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.* 
```{r}

```
*Add some text here showing the model output returning paramaters.*

>>>>

---
title: "A sample workflow"
authors: "Lizzie, Victor"
date: "April 2025"
date-format: "MMMM YYYY"
toc: true
number-sections: true
highlight: pygments
execute:
  enabled: true
format:
  html:
    html-math-method: katex
    theme:
      - lux
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
    page-layout: full
    fig-width: 8
    fig-height: 4
  pdf:
    keep-tex: true
    fig-width: 5.5
    fig-height: 5.5
    code-overflow: wrap
    monofontoptions:
      - Scale=0.5
knitr:
  opts_chunk:
    comment: ''
  opts_knit:
    global.par: TRUE
format-links: false
---

@@

**Note**: data from Jonathan comes from *LivingPlanetIndexDatabase_2025-02-01_19.19.51*.

@@

```{r}
set.seed(20012029)

library(ggplot2)

lpi_subset <- read.table("/home/victor/projects/forecastflows/analyses/jdavies/African_mammals.txt")

```

@@

What is trend in mammal species population numbers in Africa, both overall and how does it vary acrosss species and populations? 
**Note**: we will actually focus on African mammals with 10+ populations with 5+ observations per population.

### Simulating data
Now we start to simulate!
*Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.* 
```{r}

nspecies <- 9
npops_persp <- round(runif(nspecies, 1, 3)) # no. of different populations per species
npops <- sum(npops_persp) # total no. of different populations
nobs_perpop <- round(runif(npops, 5, 15)) # no. of different observations per population

spid_perpop <- rep(1:nspecies, times = npops_persp)
spid_perobs <- rep(spid_perpop, times = nobs_perpop)
popid_perobs <- rep(1:npops, times = nobs_perpop)

years <- c()
for(np in nobs_perpop){
  years_p <- sample(1960:2020, size = np, replace = FALSE)
  years <- c(years, years_p)
}

# nested intercepts
mu_species_alpha <- rnorm(n = nspecies, mean = 200, sd = 25)
sig_species_alpha <- rnorm(n = nspecies, mean = 5, sd = 2)
mu_pop_alpha <- rnorm(n = npops, mean = mu_species_alpha[spid_perpop])
sig_pop_alpha <- abs(rnorm(n = npops))
alpha <- rnorm(n = npops, mean = mu_pop_alpha, sd = sig_pop_alpha)

# nested slopes
mu_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.1)
sig_species_beta <- rnorm(n = nspecies, mean =0, sd = 0.5)
mu_pop_beta <- rnorm(n = npops, mean = mu_species_beta[spid_perpop], sd = 0.2)
sig_pop_beta <- abs(rnorm(n = npops))
beta <- rnorm(n = npops, mean = mu_pop_beta, sd = sig_pop_beta)

sigma_obs <- abs(rnorm(1, mean = 10, sd = 1))

y <- c()
for(p in 1:npops){
  nobs_p <- nobs_perpop[p]
  alpha_p <- alpha[p]
  beta_p <- beta[p]

  years_p <- years[popid_perobs == p]

  yhat_p <- alpha_p + beta_p * (years_p - 1980)
  # yhat_p <- alpha_p
  y_p <- rnorm(n = nobs_p, mean = yhat_p, sd = sigma_obs)
  y <- c(y, y_p)
}

dataplot <- data.frame(
  y,
  year = years - 1980,
  species = as.character(spid_perobs),
  population = as.character(popid_perobs)
)

ggplot(data = dataplot) +
  geom_line(aes(x = year+1980, y = y, 
                group = paste0(species, population),
                color = species)) +
  theme_bw()

nestlm <- lm(y ~ year * species/population, data = dataplot)
coefs <- summary(nestlm)$coefficients

species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("species", sp), "Std. Error"]}))

species_slope <- c(coefs["year", "Estimate"], sapply(2:nspecies, function(sp){coefs["year", "Estimate"] + coefs[paste0("year:species", sp), "Estimate"]}))
species_slope_sd <- c(coefs["year", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("year:species", sp), "Std. Error"]}))

intercepts <- data.frame(obs=mu_species_alpha, est=species_intercept, est.sd=species_intercept_sd, species = as.character(1:nspecies))
ggplot(data = intercepts) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
  geom_pointrange(aes(ymin = est-est.sd, ymax = est+est.sd, x = obs, y = est, color = species)) +
  theme_bw()

slopes <- data.frame(obs=mu_species_beta, est=species_slope, est.sd=species_slope_sd, species = as.character(1:nspecies))
ggplot(data = slopes) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
  geom_pointrange(aes(ymin = est-est.sd, ymax = est+est.sd, x = obs, y = est, color = species)) +
  theme_bw()

```
*Add some text here showing the model output returning parameters.*
